R for Clinical Study Reports and Submission

### 2021-09-15 index.Rmd
update merck.com link
update index.Rmd
fix typo

Welcome to R for clinical study report and submission. 
Clinical study reports (CSR) are the essential part of clinical trials development. 
a CSR is an "integrated" full scientific report of an individual clinical trial.

The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
provide guidance to assist sponsors in the development of a CSR. 
In this book, you will learn how to use R to prepare CSR and submit to regulatory agency.

This is the work-in-progress draft of the book.

## List of Authors and Contributors

@@

- Author: contribute majority content of at least one chapter. 
- Contributor: [contribute at least one commit to the source code](https://stash.merck.com/projects/BARDS/repos/r4csr/browse)

>>>>

Welcome to R for clinical study report and submission. 
Clinical study reports (CSR) are essential parts of clinical trials development. 
a CSR is an "integrated" full scientific report of an individual clinical trial.

The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
provide guidance to assist sponsors in the development of a CSR. 
In this book, you will learn how to use R to prepare CSR and submit to regulatory agencies.

**This is the work-in-progress draft of the book.**

## List of Authors and Contributors {-}

@@

- Author: contribute majority content of at least one chapter. 
- Contributor: [contribute at least one commit to the source code](https://github.com/elong0527/r4csr)

### 2021-09-15 intro.Rmd
fix typo

In clinical trial development, developers needs to organize source code to deliver Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets and table, listing and figures (TLFs).
For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
A consistent and well-defined folder structure is crucial to manage a clinical trial analysis and reporting (A&R) project.

We recommend to use R package folder structure to organize all the A&R related source code and documentation for a clinical trial A&R project. 
By using the R package folder structure, you can follow the convention to organize your code as other contributors on CRAN.
This consistent approach simplifies the communication for all developers within and across organizations.

- For a new R developer, it is an essential step to learn R package when you want to share your work with others. 
You will learn one folder structure that is widely used in R community with outstanding tutorial and tools for free.

- For an experienced R developer, there is minimal learning curve.

- For an organization, it simplify process, tool, template, and training development, because a unified folder structure is used to develop and maintain standard tool and analysis project.

@@

In addition, R package folder structure are also recommended to develop Shiny app as discussed in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.

@@

This book is an intermediate level book by assuming reader has equipped some R programming and clinical development knowledge.
The assumption of each part is as below:

- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
We expect readers are familiar with data manipulation in R. Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/), 
[R for Data Science](https://r4ds.had.co.nz/) and [Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).

@@

- "Anything that can be automated, should be automated." 
- "Do as little as possible by hand. Do as much as possible with functions." 
- "The goal is to spend your time thinking about what you want to do 
rather than thinking about the minutiae of package structure."

>>>>

In clinical trial development, developers need to organize source code to deliver 
Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets 
and tables, listings, and figures (TLFs).
For example, a Phase 3 clinical trial typically requires hundreds of TLFs. 
A consistent and well-defined folder structure is crucial 
to managing a clinical trial analysis and reporting (A&R) project.

We recommend using R package folder structure to organize all the A&R related 
source code and documentation for a clinical trial A&R project. 
By using the R package folder structure, 
you can follow the convention to organize your code as other contributors on CRAN.
This consistent approach simplifies the communication for all developers 
within and across organizations.

- For a new R developer, 
it is an essential step to develop R packages 
when you want to share your work with others. 
You will learn one folder structure that is widely used 
in the R community with outstanding tutorials and tools for free.

- For an experienced R developer, there is a minimal learning curve.

- For an organization, it simplifies the process, tool, template, 
and training development, because a unified folder structure is used 
to develop and maintain standard tool and analysis projects.

@@

In addition, R package folder structure is also recommended 
to develop Shiny app as discussed 
in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book 
and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.

@@

This book is an intermediate-level book by 
assuming readers have R programming and clinical development knowledge.
The assumption of each part is as below:

- Part 1, "Delivering TLFs in CSR" provides general information with examples 
to create tables, listings, and figures. 
In this part, we assume readers are individual contributors to a clinical project
with some experience in R programming. 
We expect readers are familiar with data manipulation in R. 
Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
[R for Data Science](https://r4ds.had.co.nz/) and 
[Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).

@@

- "Anything that can be automated, should be automated." 
- "Do as little as possible by hand. Do as much as possible with functions." 
- "The goal is to spend your time thinking about what you want to do 
rather than thinking about the minutiae of the package structure."

### 2021-09-15 tlf-overview.Rmd
fix typo

In clinical trial development, it is a critical step to submit information on medicinal products to regulatory agencies. [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
has become a worldwide regulatory submission standard format. 
For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd). Clinical Data Interchange Standards Consortium (CDISC) provided a 
[pilot project
following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).

@@

A typical CSR contain full details of an individual clinical study 
including clinical and statistical description, presentations, and analyses. 
A large number of tables and figures are incorporated into the main text and appendices 
of a CSR. In CDISC pilot project, an [example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf) 
is also provided. If you are interested in more examples of clinical study reports, 
the European Medicines Agency (EMA) publishes clinical data including 
clinical study reports submitted by pharmaceutical companies 
on [EMA clincal data website](https://clinicaldata.ema.europa.eu/web/cdp/home).

Building CSR is a team work with clinicians, medical writers, statisticians and statistical programmers.
Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
In an organization, a group of statisticians and statistical programmers commonly work together to 
define, develop, validate and deliver tables, listings and figures (TLFs) required for a CSR.
Microsoft Word is widely used to prepare CSR in pharmaceutical industry. 
Therefore, `rtf`, `doc`, `docx` are commonly used format to deliver TLFs.

In this chapter, our focus is to illustrate how to create tables, listings and figures (TLFs) in RTF format 
that is commonly used in a CSR. The examples are in compliance of 
[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)

> FDA's PDF specification is a general reference, each organization may define 
> more specific TLF format requirement that can be different from the examples in this book.

@@

- Section 10: Study patients 
- Section 11: Efficacy evaluation
- Section 12: Safety evaluation 
- Section 14: Tables, Figures and Graphs referred to but not included in the text
- Section 16: Appendices

@@

We used public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

@@

The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

@@

In this part, we mainly use R packages below to illustrate 
how to deliver TLFs in CSR.

@@

`tidyverse` is a collection of R packages to simplify the workflow to manipulate, 
visualize and analyze data in R. 
Those R packages share [the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
and easy to use for interactive data analysis.

@@

`r2rtf` is an R package to create production ready tables and figures in RTF format.
The R package is designed to

- provide simple "verb" functions that correspond to each component of a table, 
  to help you translate data frame to table in RTF file.
- enable pipes (`%>%`). 
- only focus on **table format**. 
  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)

@@

Other extended examples and features are covered in 
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

@@

Below is the meaning of relevant variables. 
More information can be found in help page of the dataset (`?r2rtf_adae`)

@@

```{r}
r2rtf_adae %>%
  select(USUBJID, TRTA, AEDECOD) %>%
  head(4)
```
`dplyr` and `tidyr` packages within `tidyverse` are used 
for data manipulation to create a data frame 
that contains all the information we want to add in an RTF table.

@@

All these verbs is designed to enables usage of pipes (`%>%`).
A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)

@@

In this example, the input of `col_rel_width` is a vector with same length for number of columns. 
This argument defines the relative width of each column 
within a pre-defined total column width.

@@

In previous example, we found issue of misaligned column header. 
We can fix the issue by using `rtf_colheader` function.

In `rtf_colheader`, `colheader` argument is used to provide content of column header. 
We use `"|"` to separate the columns.

@@

Below is an example to make first column left aligned and center aligned for the rest.

@@

If we want to remove the top border of `"Adverse Events"` in header, 
we can change default value `"single"` to `""` in `border_top` argument as showed below.

>>>>

In clinical trial development, it is a critical step to 
submit information on medical products to regulatory agencies. 
[Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
has become a worldwide regulatory submission standard format. 
For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd). Clinical Data Interchange Standards Consortium (CDISC) provided a 
[pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).

@@

A typical CSR contain full details of an individual clinical study 
including clinical and statistical description, presentations, and analyses. 
A large number of tables and figures are incorporated into the main text and appendices of a CSR. 
In CDISC pilot project, an 
[example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf) 
is also provided. If you are interested in more examples of clinical study reports, 
the European Medicines Agency (EMA) publishes clinical data including 
clinical study reports submitted by pharmaceutical companies 
on [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).

Building CSR is teamwork with clinicians, medical writers, statisticians, and statistical programmers. 
Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
In an organization, a group of statisticians and statistical programmers commonly work together to 
define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR.
Microsoft Word is widely used to prepare CSR in the pharmaceutical industry. 
Therefore, `rtf`, `doc`, `docx` are commonly used formats to deliver TLFs.

In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in the RTF format 
that is commonly used in a CSR. The examples are in compliance of 
[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)

> FDA's PDF specification is a general reference, each organization can define 
> more specific TLF format requirement that can be different from the examples in this book.

@@

- Section 10: Study patients 
- Section 11: Efficacy evaluation
- Section 12: Safety evaluation 
- Section 14: ables, Figures, and Graphs referred to but not included in the text
- Section 16: Appendices

@@

We used publicly available CDISC pilot 
[study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

@@

The dataset structure follows 
[CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

@@

In this part, we mainly use the R packages below to illustrate 
how to deliver TLFs in CSR.

@@

`tidyverse` is a collection of R packages to simplify the workflow to manipulate, 
visualize and analyze data in R. 
Those R packages share 
[the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
and are easy to use for interactive data analysis.

@@

`r2rtf` is an R package to create production-ready tables and figures in RTF format.
The R package is designed to

- provide simple "verb" functions that correspond to each component of a table, 
  to help you translate data frame to a table in RTF file.
- enable pipes (`%>%`). 
- only focus on **table format**. 
  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)

@@

Other extended examples and features are covered in the 
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

@@

Below is the meaning of relevant variables. 
More information can be found in the help page of the dataset (`?r2rtf_adae`)

@@

```{r}
r2rtf_adae %>%
  select(USUBJID, TRTA, AEDECOD) %>%
  head(4)
```

`dplyr` and `tidyr` packages within `tidyverse` are used 
for data manipulation to create a data frame 
that contains all the information we want to add in an RTF table.

@@

All these verbs are designed to enable the usage of pipes (`%>%`).
A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)

@@

In this example, the input of `col_rel_width` is a vector 
with the same length for the number of columns. 
This argument defines the relative width of each column 
within a pre-defined total column width.

@@

In the previous example, we found the issue of a misaligned column header. 
We can fix the issue by using the `rtf_colheader` function.

In `rtf_colheader`, `colheader` argument is used to provide the content of the column header. 
We use `"|"` to separate the columns.

@@

Below is an example to make the first column left-aligned and center-aligned for the rest.

@@

If we want to remove the top border of `"Adverse Events"` in the header, 
we can change default value `"single"` to `""` in `border_top` argument as showed below.
